CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-09-21 docs/CrystalPlan_paper/paper.tex
Refs #182: Had to use .ps images. Added a screenshot
Refs #182: Extra screenshot.

\documentclass{iucr}              % DO NOT DELETE THIS LINE

@@

\usepackage[pdftex]{graphicx}

@@

\subsubsection{Main Program Workflow.}
Some screenshots of the GUI are shown in Figure \ref{fig:screenshots}. The main
window consists of a series of tabs, with workflow proceeding from left to right:

\begin{itemize}
  \item Q-Space tab: define the volume of interest (minimum d-spacing) and
resolution of reciprocal space. Detectors tab: load instrument detector geometry
files, and enable or disable individual detectors. Goniometer tab: to choose the
goniometer and degrees of orientation freedom of the instrument.

  \item Sample tab: enter or load the crystal's lattice parameters and UB matrix.

  \item Try an Orientation tab: interactively rotate a sample and observe the
  changes in coverage in real-time.

  \item Add Orientations tab: type in a list of sample orientation angles to add
  to the experiment plan.

@@

\subsubsection{Reciprocal Space Coverage Visualization.}
Figure \ref{fig:volume_view}a) shows a screenshot of the reciprocal space
coverage 3D interface. The 3D view can represent q-space in two ways: volume coverage, or single-peak
coverage.

@@

The reflections view (Figure \ref{fig:refl_view}a) displays each hkl as a
small sphere; the color of which indicates the number of times the hkl is predicted to be measured
in the current experiment plan.

Both views can take into crystal symmetry in their display (Figures
\ref{fig:volume_view}b and \ref{fig:refl_view}b), using the techniques described
previously.

Finally, both views display quick coverage statistics, showing the percentage of
volume (or of the number of peaks) measured, as well as the proportion of
redundancy (voxels or peaks measured 2 or more times).

@@

Live data processing    …  interface with ISAW … InelasCasetic instruments ??

@@

\begin{figure}
\caption{Ewald sphere construction showing reciprocal-space detector coverage
for a wide-bandwidth Laue diffractometer. The smaller blue circle represents the
$1/\lambda_{max}$ limit of measurement, with the larger green circle
representing $1/\lambda_{min}$. The origin of reciprocal space (000) is common to both
spheres, but the sample position in the construction (S1 or S2) depends on
$\lambda$. The angular coverage of the detector is represented by the dashed
wedge, and projects onto each sphere of radius $1/\lambda$ an arc. The red wedge
between the two $\lambda$ is built by projecting straight lines between the
$q_{min}$ and $q_{max}$ limits, giving the volume in reciprocal space that is
measured at a single sample oriention. }
%\includegraphics[bb=0in 0in 2in 2in, width=2in]{ewald_fig.pdf} 
\label{fig:ewald_fig} 
\end{figure}

\begin{figure}
\caption{Screenshots of the main GUI}
%\includegraphics{fig1.ps}
\label{fig:screenshots}
\end{figure}

@@

>>>>

\documentclass[final]{iucr}              % DO NOT DELETE THIS LINE

@@

\usepackage{graphicx}
%\usepackage[pdftex]{graphicx}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

@@

\subsubsection{Main Program Workflow.}
A screenshot of the main window GUI is shown in Figure \ref{fig:screenshots}.
The main window consists of a series of tabs, with workflow proceeding
intuitively from left to right:

\begin{itemize}
  	\item Q-Space tab: define the volume of interest (minimum d-spacing) and
resolution of reciprocal space. 
 	\item Detectors tab: load instrument detector geometry
files, and enable or disable individual detectors. 
	\item Goniometer tab: to choose the
goniometer and degrees of orientation freedom of the instrument.
  	\item Sample tab: enter or load the crystal's lattice parameters and UB
  matrix.
  	\item Try an Orientation tab: interactively rotate a sample and observe the
  changes in coverage in real-time.
  	\item Add Orientations tab: type in a list of sample orientation angles to
  add to the experiment plan.

@@

\subsubsection{Reciprocal Space Coverage Visualization.}
Figure \ref{fig:volume_view}a) shows a screenshot of the reciprocal space
coverage 3D interface. The 3D view can represent q-space in two ways: volume coverage, or single-peak
coverage. The view shown can be rotated, translated or zoomed using the mouse or
keyboard.

@@

The reflections view (Figure \ref{fig:refl_view}a) displays each $hkl$ as a
small sphere; the color of which indicates the number of times the $hkl$ is
predicted to be measured in the current experiment plan.

Both views can take into crystal symmetry in their display using the techniques
described previously. 
The volume coverage (Figures \ref{fig:volume_view}b) fills in the sphere of
coverage; 
whereas the single reflection view (Figure \ref{fig:refl_view}b) leaves only
the primary reflections (one eight of the total, in this example).

Finally, both views display quick coverage statistics in the lower right
corner, showing the percentage of volume (or of the number of peaks) measured,
as well as the proportion of redundancy (voxels or peaks measured 2 or more
times).

\subsubsection{Single Reflection View.}
Figure \ref{fig:single_reflection} shows a screenshot of the GUI allowing one to
observe in detail the predicted measurements of a particular peak, by clicking
in the 3D view or by typing in the $hkl$ values. The position on the detector
and wavelength of measurement are displayed. An additional window (not shown) allows the user to place the peak on a particular
position on a detector by automatically finding the necessary sample orientation
angles. This is used, for example, to center important peaks in the most
sensitive part of a detector.

@@

We have described CrystalPlan, a software program designed to quickly produce
efficient experiment plans for wide bandwidth time-of-flight Laue
diffractometers. This program will help maximize scientific productivity at SNS
beamlines such as TOPAZ.

One planned future development for CrystalPlan include extending the software
to help predict coverage in inelastic neutron scattering experiments. Another
desired feature is the introduction of feedback loops to help determine the
right time to end data acquisition at a particular sample orientation.
CrystalPlan would interface with data analysis software such as ISAW and use
recently developped capabilities for real-time data analysis to monitor
the  signal to noise ratio on particular peaks. When the signal becomes
sufficient, the sample will automatically be moved to the next sample
orientation, further minimizing data acquisition time. Combined with advanced
sample changers coming on-line at TOPAZ, this will allow for extremely
efficient, unattended data acquisition.

@@

\begin{figure}
\caption{Ewald sphere construction showing reciprocal-space detector coverage
for a wide-bandwidth Laue diffractometer. The smaller blue circle represents the
$1/\lambda_{max}$ limit of measurement, with the larger green circle
representing $1/\lambda_{min}$. The origin of reciprocal space (000) is common to both
spheres, but the sample position in the construction (S1 or S2) depends on
$\lambda$. The angular coverage of the detector is represented by the dashed
wedge, and projects onto each sphere of radius $1/\lambda$ an arc. The red wedge
between the two $\lambda$ is built by projecting straight lines between the
$q_{min}$ and $q_{max}$ limits, giving the volume in reciprocal space that is
measured at a single sample oriention. }

%\includegraphics[bb=0in 0in 8in 6in, scale=0.20]{ewald_fig.pdf}
\includegraphics[width=3.5in]{ewald_fig.ps}

\label{fig:ewald_fig} 
\end{figure}

\begin{figure}
\caption{Screenshot of the main GUI window. Six more interface windows (not
shown) are presented to the user when he or she clicks on the tabs at the
top of the main window.}
\includegraphics{main_gui.eps}
\label{fig:screenshots}
\end{figure}

@@

\begin{figure}
\caption{GUI showing details of the predicted measurement positions for a
single $hkl$.}
\includegraphics[width=2.5in]{single_reflection_2.eps}
\label{fig:single_reflection}
\end{figure}
